Tests for Measurement Invariance When Subgroups Are Unknown

### 2012-07-03 imps2012/invariance.tex
some small touch-ups, changed definition of efp to match the Psychometrika paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@144 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\title[Measurement Invariance]{Tests for Measurement Invariance when Subgroups are Unknown}
\author{Ed Merkle\inst{1} \and Achim Zeileis\inst{2}}
\institute{\inst{1} University of Missouri \and \inst{2} Universit\"{a}t Innsbruck}
\date{Supported by grant SES-1061334 from the National
  Science Foundation}

@@

\section{Tests Along a Continuum}
% \frame{
%   \frametitle{Proposed Tests}
%   \begin{itemize}
%     \item In contrast to existing tests of measurement invariance, the
%       proposed tests offer the abilities to:
%       \begin{itemize}
%         \item Test for measurement invariance when groups are
%           ill-defined (e.g., when the grouping variable is continuous).
%         \item Test for measurement invariance in any subset of model
%           parameters.
%         \item Interpret the nature of measurement invariance violations.
%       \end{itemize}
%   \end{itemize}}

\frame{
  \frametitle{Proposed Tests}
  \begin{itemize}
    \item We propose a number of tests for measurement invariance
      along a continuous auxiliary variable.
    \item Tests are based on individual terms ({\em scores}) of the gradient.  The scores tell us how well a particular parameter describes a particular individual.
  % Equation of gradient vs scores for individuals
  \end{itemize}
\begin{eqnarray*}
  \displaystyle\sum_{i=1}^{n} s(\hat{\bm{\theta}} ; \bm{x}_i) =
  {\bm 0},\text{ where} \\
 \\
  s(\hat{{\bm \theta}}; \bm{x}_i) = \frac{\partial}{\partial \bm{\theta}} \log
\text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}
\end{eqnarray*}}

@@

\begin{frame}[fragile]
  \frametitle{Aggregating Scores}
 % define cumulative scores and other notation
  \begin{itemize}
    \item We need a way to aggregate scores across people so that we
      can draw some general conclusions.
      \begin{itemize}
        \item Order individuals by the auxiliary variable.\\ \ \\
        \item Define $t \in (1/n, n)$.
          The {\em empirical cumulative score process} is defined by:\\
  \begin{equation*}
    \bm{B}(\hat{\bm{\theta}}; t) = \frac{1}{\sqrt{n}}
    \displaystyle\sum_{i=1}^{\lfloor nt 
      \rfloor} s(\hat{\bm{\theta}} ; \bm{x}_i).
    \end{equation*}
  where $\lfloor nt \rfloor$ is the integer part of $nt$.
      \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Tests}
 % show that cumulative scores follow a Brownian bridge under H_0
  \begin{itemize}
    \item Under the hypothesis of
      measurement invariance, a functional central limit
      theorem holds:
  \begin{equation*}
  {\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{B}}(\widehat{{\bm{\theta}}}; \cdot) 
\overset{d}{\rightarrow} {\bm B}^{0}(\cdot),
  \end{equation*}
 where ${\bm{I}}(\widehat{{\bm{\theta}}})$ is the observed
 information matrix and ${\bm B}^{0}(\cdot)$ is a
 $p$-dimensional Brownian bridge.\\ \ \\

@@

\section{Tests Along an Ordinal Variable}
\frame{
  \frametitle{Tests Along Ordinal Variables}
  \begin{itemize}
    \item The statistics described above do not immediately translate
      to ordinal auxiliary variables, where there is no unique ordering of
      individuals.
    \item To obtain a test statistic, we allow all individuals with the
      same value of the 
      auxiliary variable to simultaneously enter into the cumulative
      sum.  We then take the maximum of the cumulative sum.
    \item Critical values are obtained by summing bins of a Brownian
      bridge, where bin sizes match the observed bin sizes of the 
      ordinal variable.
  \end{itemize}}

>>>>

\title[Measurement Invariance]{Tests for Measurement Invariance When Subgroups Are Unknown}
\author{Ed Merkle\inst{1} \and Achim Zeileis\inst{2}}
\institute{\inst{1} University of Missouri \and \inst{2} Universit\"{a}t Innsbruck}
\date{Supported by grant SES-1061334 from the National
  Science Foundation}

@@

\section{Tests along a Continuum}
% \frame{
%   \frametitle{Proposed Tests}
%   \begin{itemize}
%     \item In contrast to existing tests of measurement invariance, the
%       proposed tests offer the abilities to:
%       \begin{itemize}
%         \item Test for measurement invariance when groups are
%           ill-defined (e.g., when the grouping variable is continuous).
%         \item Test for measurement invariance in any subset of model
%           parameters.
%         \item Interpret the nature of measurement invariance violations.
%       \end{itemize}
%   \end{itemize}}

\frame{
  \frametitle{Proposed Tests}
  \begin{itemize}
    \item We propose a number of tests for measurement invariance
      along a continuous auxiliary variable.
    \item Tests are based on individual terms of the gradient ({\em scores}).  The scores tell us how well a particular parameter describes a particular individual.
  % Equation of gradient vs scores for individuals
  \end{itemize}
\begin{eqnarray*}
  \displaystyle\sum_{i=1}^{n} s(\hat{\bm{\theta}} ; \bm{x}_i) =
  {\bm 0},\text{ where} \\
 \\
  s(\hat{{\bm \theta}}; \bm{x}_i) = \frac{\partial}{\partial \bm{\theta}} \log
\text{L}({\bm{x}}_i, {\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}
\end{eqnarray*}}

@@

\begin{frame}[fragile]
  \frametitle{Aggregating Scores}
 % define cumulative scores and other notation
  \begin{itemize}
    \item We need a way to aggregate scores across people so that we
      can draw some general conclusions.
      \begin{itemize}
        \item Order individuals by the auxiliary variable.\\ \ \\
        \item Define $t \in (0, 1)$.
          The {\em empirical cumulative score process} is defined by:\\
  \begin{equation*}
    \bm{B}(\hat{\bm{\theta}}; t) = 
      {\bm{\widehat I}}^{-1/2} \frac{1}{\sqrt{n}}
    \displaystyle\sum_{i=1}^{\lfloor nt 
      \rfloor} s(\hat{\bm{\theta}} ; \bm{x}_i).
    \end{equation*}
  where $\lfloor nt \rfloor$ is the integer part of $nt$
  and $\bm{\widehat I}$ is some consistent covariance matrix estimate, e.g.,
  the observed information matrix ${\bm{I}}(\widehat{{\bm{\theta}}})$.

      \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Tests}
 % show that cumulative scores follow a Brownian bridge under H_0
  \begin{itemize}
    \item Under the hypothesis of
      measurement invariance, a functional central limit
      theorem holds:
  \begin{equation*}
  {\bm{B}}(\widehat{{\bm{\theta}}}; \cdot) \overset{d}{\rightarrow} {\bm B}^{0}(\cdot),
  \end{equation*}
 where ${\bm B}^{0}(\cdot)$ is a $p$-dimensional Brownian bridge.\\ \ \\

@@

\section{Tests along an Ordinal Variable}
\frame{
  \frametitle{Tests along Ordinal Variables}
  \begin{itemize}
    \item The statistics described above do not immediately translate
      to ordinal auxiliary variables, where there is no unique ordering of
      all individuals.
    \item To obtain a test statistic, we allow all individuals with the
      same value of the 
      auxiliary variable to simultaneously enter into the cumulative
      sum.  We then take the maximum of the cumulative sum.
    \item Critical values are obtained by summing bins of a Brownian
      bridge, where bin sizes match the observed bin sizes of the 
      ordinal variable.
  \end{itemize}}
